Towards multiversal modal operators for homotopy type theory

### 2023-11-29 slides-hottest2023.tex
Sequences provided by LEM and sequences depending on an environment

\usepackage[normalem]{ulem}
\usepackage{mathtools}
\usepackage{booktabs}
\usepackage{stmaryrd}
\usepackage{amssymb}
\usepackage{manfnt}
\usepackage{array}
\usepackage{ragged2e}
\usepackage{multicol}
\usepackage{tabto}
\usepackage{xstring}
\usepackage{proof}
\usepackage[all]{xy}
\xyoption{rotate}
\usepackage{tikz}
\usetikzlibrary{calc,shapes,shapes.callouts,shapes.arrows,patterns,fit,backgrounds,decorations.pathmorphing,positioning,svg.path}
\hypersetup{colorlinks=true}

@@

@@

  \jnote{7-}{
    The class of well preorders is stable under cartesian products, lists and
    trees, by Dickson's Lemma, Higson's Lemma and Kruskal's Theorem,
    respectively. However, in their naive formulations, these are merely
    theorems of classical mathematics. For general constructive results, the
    definition of ``well'' needs to be improved.

@@

  \textbf{Thm.} Every sequence~$\alpha : \NN \to \NN$ is \hil{good} in that
  there exist~$i < j$ with~$\alpha(i) \leq \alpha(j)$.
  \pause

  \emph{Proof.} \emph{(offensive?)} By~\badbox{\textsc{lem}}, there is a
  minimum~$\alpha(i)$.
  Set~$j \defeq i + 1$. \qed\par
  \pause

  \textbf{Def.} (classically) A preorder~$X$ is \hil{well} iff every sequence~$\NN \to X$ is good.

@@

  \textbf{Def.} $\displaystyle\mathsf{Well}(X,{\leq}) \defeq
    \prod\limits_{\alpha \? \NN \to X} \, \Bigl\|\sum\limits_{i \? \NN} \sum\limits_{j \? \NN}
      i < j \times \alpha(i) \leq \alpha(j)\Bigr\|_{-1}$
  \pause
  \qquad\scalebox{2}{\dbend}
  \medskip

@@

>>>>

\usepackage[normalem]{ulem}
\usepackage{mathtools}
\usepackage{booktabs}
\usepackage{stmaryrd}
\usepackage{amssymb}
\usepackage{manfnt}
\usepackage{array}
\usepackage{ragged2e}
\usepackage{multicol}
\usepackage{tabto}
\usepackage{xstring}
\usepackage{proof}
\usepackage{agda}
\usepackage[all]{xy}
\xyoption{rotate}
\usepackage{tikz}
\usetikzlibrary{calc,shapes,shapes.callouts,shapes.arrows,patterns,fit,backgrounds,decorations.pathmorphing,positioning,svg.path}
\hypersetup{colorlinks=true}

@@

\newcommand{\dnote}[1]{%
  \begin{tabular}{@{}m{2em}@{}m{0.80\textwidth}@{}}%
    \textdbend &#1%
  \end{tabular}%
  \par
}

@@

  \jnote{7-}{
    The class of well quasiorders is stable under cartesian products, lists and
    trees, by Dickson's Lemma, Higson's Lemma and Kruskal's Theorem,
    respectively. However, in their naive formulations, these are merely
    theorems of classical mathematics. For general constructive results, the
    definition of ``well'' needs to be improved.

@@

  \begin{block}{}
    \justifying
    \textbf{Thm.} Every sequence~$\alpha : \NN \to \NN$ is \hil{good} in that
    there exist~$i < j$ with~$\alpha\,i \leq \alpha\,j$.
  \end{block}
  \pause
  \vspace*{-0.6em}
  \emph{Proof.} \emph{(offensive?)} By~\badbox{\textsc{lem}}, there is a
  minimum~$\alpha\,i$.
  Set~$j \defeq i + 1$. \qed\par
  \pause

  \textbf{Def.} (classically) A quasiorder~$X$ is \hil{well} iff every sequence~$\NN \to X$ is good.

@@

  \textbf{Def.} $\displaystyle\mathsf{Well}(X,{\leq}) \defeq
    \prod\limits_{\alpha \? \NN \to X} \, \Bigl\|\sum\limits_{i \? \NN} \sum\limits_{j \? \NN}
      i < j \times \alpha\,i \leq \alpha\,j\Bigr\|_{-1}$
  \pause
  \qquad\scalebox{2}{\dbend}
  \medskip

@@

\begin{frame}{Sequences provided by \textsc{lem}}
  \justifying
  \begin{block}{}
    \justifying
    \textbf{Lemma.} Let~$X$ be a well quasiorder. Let~$\alpha : \NN \to X$ be a
    sequence. Assuming~\bad{\textsc{lem}}, there merely is a monotonic
    subsequence~$\alpha\,i_0 \leq \alpha\,i_1 \leq \cdots$.
  \end{block}
  \vspace*{-0.6em}
  \emph{Proof.} The set
  \[ I \defeq \{ i \in \NN \,|\, \neg\mathop{\exists}\limits_{j \in \NN} i < j \times \alpha\,i \leq \alpha\,j
  \} \]
  cannot be in bijection with~$\NN$, as else the~$I$-extracted subsequence
  of~$\alpha$ would not be good. By~\bad{\textsc{lem}}, the set~$I$ is finite.
  Any index~$i_0$ larger than all the indices in~$I$ is a suitable starting
  point for a monotonic subsequence. \qed
  \medskip
  \pause

  \begin{block}{}
    \justifying
    \textbf{Prop.} Let~$X$ and~$Y$ be well quasiorders.
    Assuming~\bad{\textsc{lem}},~$X \times Y$ is well.
  \end{block}
  \vspace*{-0.6em}
  \emph{Proof.} Let an infinite sequence~$\gamma : \NN \to X \times Y$ be
  given. Write~$\gamma\,k = (\alpha\,k,\beta\,k)$. By the lemma, there is a
  monotonic subsequence~$\alpha\,i_0 \leq \alpha\,i_1 \leq \cdots$. Because~$Y$
  is well, there are indices~$n < m$ such that~$\beta\,i_n \leq \beta\,i_m$. As
  also~$\alpha\,i_n \leq \alpha\,i_m$, the sequence~$\gamma$ is good.
  \medskip
  \pause

  \dnote{\emph{We cannot trust~\bad{\textsc{lem}}-provided sequences to be
  available in the type~$\NN \to \NN$. Similarly with~\bad{\textsc{dc}}.}}
\end{frame}

\begin{frame}{Sequences depending on an environment}
  Let~$X$ be an hset such that there is no surjection~$\NN \twoheadrightarrow X$.
  Then the type~$\NN \to X$ misses the \hil{generic enumeration} of~$X$.
  \medskip
  \pause

  \emph{Idea:} Approximate a (fictitious) surjection~$\NN \twoheadrightarrow X$
  by finite sequences~$\sigma = x_0 \ldots x_{n-1} \? X^*$. Over time, such an
  approximation can grow to
  \begin{enumerate}
    \item one of $\sigma y$, where~$y \? X$, so it becomes \emph{more defined}; or
    \item one of~$\sigma \tau$, where~$\tau \? X^*$ such that~$a \in
    \sigma\tau$, so it becomes \emph{more surjective}.
  \end{enumerate}
  \medskip
  \pause

  Formally, we have the type~$\nabla\underline{X}$ of stage-dependent values of~$X$:
\end{frame}
